Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-10-04 MS/ms.Rmd
Major overhaul of survey section. Integrates its purpose better with the heterogenity idea, splits results up into the sections where results make sense to convey to reader ...survye data on effects now in effect size section and survey on models in the mdoel section.

Certain types of heterogeneity, however, get in the way of understanding real biological patterns and are simply a 'nuisance' to interpreting effects. In other words, if we compare different varieties of 'apples', is it still fair to synthesize these studies to understand overall effects? In comparative physiology, this situation might be analogous to a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Here, we focus on meta-analytic tools to circumvent these issues and demonstrate that the ‘apples and oranges’ problem is not really a problem at all.

## **Survey of Meta-Analytic Practices in Comparative Physiology**

Before overviewing strategies to deal with nuisance heterogeneity, it is worth first understanding how comparative physiologists are currently conducting meta-analyses and other quantitative syntheses. As such, we conducted a literature survey to better understand the different types of effect sizes being used in comparative physiology, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled. After searching for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species, we asked: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to nuisance heterogeneity? 3) If so, was such variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

#### *Literature Search and Survey Data Collection*

All searches were conducted in Scopus between May 16 and 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Table S1 and S2 - Supplementary Materials). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

We included comparative studies / meta-analyses that were focused on physiological questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data, such as hormone levels, thermal physiological parameters (Critical thermal maximum (CTmax), critical thermal minimum (CTmin), thermal preferences, thermal breadth, lethal thermal tolerance limit (LT50), lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress measures and consequences (e.g., DNA damage, teleomere attrition), performance (e.g., maximum heart rates, sprint speed, swim speed) and growth rates, body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified the change in physiological traits in response to environmental variation (i.e., phenotypic plasticity). While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source in quantitative syntheses. We excluded quantitative syntheses interested in phenology, biomechanics, sexual size dimorphism (i.e., except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription, and quantitative genetics. There were also meta-analyses that focused on colour evolution which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because most plant synthesis papers were focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output, and greenhouse gas emissions).

@@

#### *Current Meta-analytic Practices in Comparative Physiology*

@@

The results of our literature survey show that quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`% of studies), followed closely by Fishers transformed correlation coefficients (*Zr*) (`r with(effects, effects[effect == "Zr","prop"])*100`% of studies), standardised mean differences (*SMD*) (`r with(effects, effects[effect == "SMD","prop"])*100`%  of studies), and log response ratios (*lnRR*) (`r with(effects, effects[effect == "lnRR","prop"])*100`%, of studies) (Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`% of effect sizes used within the study were deemed to be impacted by nuisance heterogeneity in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B), often through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

## **Opening New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with "Nuisance Heterogeneity"**

The key features of all effect sizes used in meta-analyses are that: 1) they are statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) they have some associated measure of precision (i.e., sampling variance) [@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite diverse; they can be the mean difference between two groups [e.g., difference in mean between a control and treatment group; e.g., @WuSeebacher2020], the slope of a regression [e.g., between body size and metabolic rate, @Uyeda2017], a correlation between two variables [(]e.g., between hormone levels and behaviour, @Holtman2016], or even the raw mean (or variance) itself [e.g., comparing CTmax or CTmin across species - e.g., @Sunday2014]. In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio, *$\ln{RR}$*. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, as long as all the means are measured in the same units (e.g., degrees Celsius) or can be converted to the same units, then the mean may provide the desired answer to the original question. We provide details on different effect measures and associated sample sizes in Table \@ref(tab:tablea).

@@

#### *Formalising Effect Size Metrics that Compare Changes in Variability Across Treatments in the Presence of "Nuisance Heterogeneity"*

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size, $v_{i}$ (which is known as the sampling variance for the effect). This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals), then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. Alternatively, robust variance estimators can also be used to deal with effect size non-independence [@Song2021; @Nakagawa2021]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation, such as differences in the methods applied, species differences, or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg T, how much of an overall effect does T have on offspring traits relative to control conditions?

Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much T has been applied in the treatments used in the sample of studies. The most we can say about it is that the effect is small, moderate or large given the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

@@

We would like to thank the Editorial board (Craig Franklin, Michaela Handel and Charlotte Rutledge) for inviting us to contribute to this special issue along with feedback on previous drafts. We would also like to thank Megan Head and two anonymous reviewers for helpful comments on previous drafts. DWAN, SMD, SN, ML and REO were supported by ARC Discovery Projects (DP210101152 to DWAN, DE180100202 to SMD, and DP200100367 to SN, ML and DWAN). PP and SB were supported by a UNSW Scientia PhD Scholarship.

>>>>

## **Nuisance Heterogeneity in Comparative Physiology**

Certain types of heterogeneity, however, get in the way of understanding real biological patterns and are simply a 'nuisance' to interpreting effects. In other words, if we compare different varieties of 'apples', is it still fair to synthesize these studies to understand overall effects? In comparative physiology, this situation might be analogous to a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth.

Here, we focus on meta-analytic solutions to circumvent nuisance heterogeneity showing how it can easily be overcome using a number of meta-analytic tools. Prior to overviewing the solutions, we conducted a literature survey to better understand the different types of effect sizes being used in comparative physiology, the susceptibility of studies to nuisance heterogeneity, and how these sources are currently (if at all) being controlled for in a meta-analysis. More specifically we asked: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to nuisance heterogeneity? 3) If so, was such variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used ‘weighted’ meta-analytic models?

All searches were conducted in Scopus between May 16 and 18, 2021, with a follow up serach on September 1, 2021 (see the Supplementary Information). Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Table S1 and S2 - Supplementary Materials). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

@@

@@

## **Opening New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with Nuisance Heterogeneity**

The key features of all effect sizes used in meta-analyses are that: 1) they are statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) they have some associated measure of precision (i.e., sampling variance) [@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite diverse; they can be the mean difference between two groups [e.g., difference in mean between a control and treatment group; e.g., @WuSeebacher2020], the slope of a regression [e.g., between body size and metabolic rate, @Uyeda2017], a correlation between two variables [(]e.g., between hormone levels and behaviour, @Holtman2016], or even the raw mean (or variance) itself [e.g., comparing CTmax or CTmin across species - e.g., @Sunday2014]. In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio, *$\ln{RR}$*. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, as long as all the means are measured in the same units (e.g., degrees Celsius) or can be converted to the same units, then the mean may provide the desired answer to the original question.

The results of our literature survey show that quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`% of studies), followed closely by Fishers transformed correlation coefficients (*Zr*) (`r with(effects, effects[effect == "Zr","prop"])*100`% of studies), standardised mean differences (*SMD*) (`r with(effects, effects[effect == "SMD","prop"])*100`%  of studies), and log response ratios (*lnRR*) (`r with(effects, effects[effect == "lnRR","prop"])*100`%, of studies) (Figure \@ref(fig:figsurv)A). We provide details on different effect measures and associated sample sizes in Table \@ref(tab:tablea). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`% of effect sizes used within the study were deemed to be impacted by nuisance heterogeneity in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B), often through meta-regression modelling.

@@

@@

#### *Formalising Effect Size Metrics that Compare Changes in Variability Across Treatments in the Presence of Nuisance Heterogeneity*

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size, $v_{i}$ (which is known as the sampling variance for the effect).

This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Weighted meta-analytic models are important to use because they 1) improve precision on meta-analytic mean estimates by accounting for sampling variance; 2) allow for a formal analysis of heterogeneity; 3) and give precedence to higher quality studies (see below). From our survey, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals), then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. Alternatively, robust variance estimators can also be used to deal with effect size non-independence [@Song2021; @Nakagawa2021]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation, such as differences in the methods applied, species differences, or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below.

Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg T, how much of an overall effect does T have on offspring traits relative to control conditions? Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much T has been applied in the treatments used in the sample of studies. The most we can say about it is that the effect is small, moderate or large given the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

@@

We would like to thank the Editorial board (Craig Franklin, Michaela Handel and Charlotte Rutledge) for inviting us to contribute to this special issue along with feedback on previous drafts. We would also like to thank Megan Head and two anonymous reviewers for very helpful comments on previous drafts. DWAN, SMD, SN, ML and REO were supported by ARC Discovery Projects (DP210101152 to DWAN, DE180100202 to SMD, and DP200100367 to SN, ML and DWAN). PP and SB were supported by a UNSW Scientia PhD Scholarship.
